% Document class for LME theses: lmedoc
% %LANGUAGE
% %CONFIG
%    The option "german" uses german.sty
%    For english papers, use the "english" option
% Possible types of theses:
% bt - Bachelor's thesis
% mt - Master's thesis
% diss - Dissertation
% sa - Student thesis
% fp - Forschungspraktikum
% pa - Projektarbeit
%\documentclass[german,bt]{lmedoc/lmedoc}
\documentclass[english,bt]{lmedoc/lmedoc}
%\documentclass[english,fp]{lmedoc/lmedoc}


%%%%%%%%%%%%%%%%%%
% pdflatex and lualatex are supported
% ++ "Umlaut" support
%    The package "inputenc" can be used to write Umlaute or the german double s
%    directly. You need to use the correct encoding, e.g. latin1.
\usepackage{iftex}
\ifPDFTeX
  \usepackage[utf8]{inputenc}
  \usepackage[T1]{fontenc}
  \usepackage{lmodern}
\else
  \ifXeTeX
     \usepackage{fontspec}
  \else 
     \usepackage{luatextra}
  \fi
\fi

%Load additional packages
\usepackage{lmedoc/packages}
% put in this file your abbreviations
\input{abbreviations.tex}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% supress messages because of underful hbox, this is not a problem
\hbadness=10000

\usepackage[backend=biber]{biblatex}

% some useful commands
\makeatletter % let's define a single dot
\DeclareRobustCommand\onedot{\futurelet\@let@token\@onedot}
\newcommand{\@onedot}{\ifx\@let@token.\else.\null\fi\xspace}
\makeatother

\usepackage{float}
\usepackage{multirow}

\usepackage{threeparttable}
\newcommand{\CellWithForcedBreak}[2][c]{
	\begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}

\usepackage[most]{tcolorbox}
\definecolor{dkgreen}{rgb}{0,0.6,0}
\definecolor{lightgray}{rgb}{0.95,0.95,0.95}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{mauve}{rgb}{0.58,0,0.82}
\definecolor{red}{rgb}{0.8,0,0}

\lstnewenvironment{Python}
{
	\lstset{frame=tb,
		language=Python,
		aboveskip=3mm,
		belowskip=0mm,
		showstringspaces=false,
		columns=flexible,
		basicstyle={\small\ttfamily},
		%		numbers=left,
		%		stepnumber=1,
		%		numberstyle=\small\color{gray},
		keywordstyle=\color{blue},
		commentstyle=\color{dkgreen},
		stringstyle=\color{red},
		breaklines=true,
		breakatwhitespace=true,
		tabsize=3}}{}

\usepackage[autostyle=true]{csquotes}
\setquotestyle[guillemets]{german} %% erzeugt quotes wie: >> xxxxx <<


% Sets the bib file
\addbibresource{literature.bib}

% When writing a large document, it is sometimes useful to work on selected sections of the document.
% Use this command to only build the document partially. Speeds up the developement cycle.
% For the final product, this has to be commented out.
%\includeonly{introduction,appendix,foo,bar}


\pagenumbering{roman}

\begin{document}
\clearpage
% %CONFIG
% This is for students' theses
  \begin{deckblatt}
    \Titel{Evaluation of Single-Head Vision Transformer (SHViT) for Volumetric Semantic Segmentation in Industrial CT Scans} % Title
    \Name{Drewes} % Last name
    \Vorname{ChangGeng} % Given name
    \Matrikelnummer{22984031}
    \Geburtsort{Yokohama, Japan} % Place of birth
    \Geburtsdatum{16.02.2000} % Date of birth
    \Betreuer{Martin Leipert} % Advisor
    \Start{18.06.2025} % Start of thesis
    \Ende{18.08.2025} % End of thesis
    %\ZweitInstitut{ZweitInstitut} % Cooperation partner
  \end{deckblatt}


\cleardoublepage
% Only necessary for thesis

Ich versichere, dass ich die Arbeit ohne fremde Hilfe und ohne Benutzung
anderer als der angegebenen Quellen angefertigt habe und dass die Arbeit
in gleicher oder "ahnlicher Form noch keiner anderen Pr"ufungsbeh"orde
vorgelegen hat und von dieser als Teil einer Pr"ufungsleistung
angenommen wurde. Alle Ausf"uhrungen, die w"ortlich oder sinngem"a"s
"ubernommen wurden, sind als solche gekennzeichnet.
\\

\vspace{2cm}

\noindent ChangGeng Drewes

\cleardoublepage


\begin{center}
\bfseries
% Abstract in German
{\selectlanguage{german}"Ubersicht}
\normalfont
\end{center}

Diese Bachelorarbeit befasst sich mit der Anpassung und Erweiterung von 2D-Vision-Transformer-Modellen, insbesondere SegFormer mit Mix-Vision Transformer (MVT) und Single-Head Vision Transfomer (SHViT), für die volumetrische 3D-Segmentierung. Der Fokus liegt auf der präzisen Segmentierung von 3D-Schuhscans. Als zentrales Qualitätsmaß wurde die F1-Metrik verwendet.

\medskip 

Die Experimente zeigen, dass die Anzahl der Convolution-Layer im SHViT-Modell erheblichen Einfluss auf die Segmentierungsgenauigkeit und den GPU-Speicherverbrauch hat. Die beste Balance wurde mit zwei Convolution-Layern im PatchEmbed-Block erzielt. Um die Speicherbelastung durch klassische Attention-Mechanismen mit $O(n^2)$-Komplexität zu verringern, wurde ein Central Self-Attention-Ansatz auf Basis eines Sliding-Window-Verfahrens implementiert. Dieser reduziert den Speicherbedarf je nach Konfiguration um bis zu 54\%, ohne die Modellleistung wesentlich zu beeinträchtigen. Zusätzlich konnten kleinere Kernelgrößen im Attention-Block die Genauigkeit weiter steigern, da damit feine Strukturen besser erfasst werden können.

\medskip

Die besten Ergebnisse wurden mit einem SHViT-Modell mit Central Self-Attention und optimierter Parametrierung erzielt, bei dem der Speicherbedarf bei unter 24\,GB GPU-Speicherbedarf liegt. Abschließend wird angeregt, sich in zukünftigen Arbeiten weiter mit der Optimierung von Attention-Mechanismen zur Reduzierung des Speicherbedarfs zu befassen und weitere Vision-Transformer-Backbones zur Verbesserung der Segmentierungsgenauigkeit auszuprobieren.

\newpage

\begin{center}
\bfseries
% Abstract in English
{\selectlanguage{english}Abstract}
\normalfont
\end{center}

This bachelor’s thesis looks at how to improve and expand 2D Vision Transformer models, particularly SegFormer with Mix-Vision Transformer (MVT) and Single-Head Vision Transformer (SHViT), for 3D segmentation of volume data. The main emphasis is on accurately segmenting 3D shoe scans. The F1-metric was used as the primary performance measure.

\medskip

Experiments show that the number of convolutional layers in the SHViT model significantly affects both segmentation accuracy and GPU memory usage. The best trade-off was achieved with two convolutional layers in the PatchEmbed block. To mitigate the high memory demands of conventional attention mechanisms with $O(n^2)$ complexity, a central self-attention approach based on a sliding-window method was implemented. Depending on the configuration, this reduces memory requirements by up to 54\% without significantly affecting model performance. Furthermore, smaller kernel sizes in the attention block led to improved accuracy, as they allowed for better capture of fine structural details.

\medskip

The best results came from an SHViT model that used central self-attention (sliding windows) with improved backbone and decoder settings, while ensuring GPU memory use stayed under 24\,GB. Finally, the thesis recommends that future work should look into improving attention mechanisms to use less memory and examine other Vision Transformer backbones to improve segmentation accuracy.

\cleardoublepage

\include{Acknowledgment}

\cleardoublepage

\tableofcontents

\cleardoublepage \pagenumbering{arabic}

\include{Introduction}   % Introduction 
\cleardoublepage
%\include{Literature_Review}   % (\chapter{})
%\cleardoublepage

\include{Methodology}   % (\chapter{})
\cleardoublepage


\include{Data_Preparation}   % (\chapter{})
\cleardoublepage
\include{Results_and_Discussion}   % (\chapter{})
\cleardoublepage

\include{Conclusion}   % Conclusion (\chapter{Conclusion}  TEXT)
%\cleardoublepage


\appendix
\cleardoublepage
\include{Appendix_WSL}   % appendix A
\include{Appendix_Code}   % appendix A
\cleardoublepage
%\include{thesis10}   % appendix B
%\cleardoublepage
%\include{thesis11}   % appendix C
%\cleardoublepage

%% Do not change, auto-generated lists of figures, tables and literature %%
% Glossar
\printunsrtglossary[type=abbreviations]
\cleardoublepage

% List of figures
\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures
\cleardoublepage

% List of tables
\addcontentsline{toc}{chapter}{\listtablename}
\listoftables
\cleardoublepage

% Literature list
% %CONFIG:
%\selectlanguage{german}{\addcontentsline{toc}{chapter}{\bibname}}
\selectlanguage{english}{\addcontentsline{toc}{chapter}{\bibname}}

\printbibliography

\end{document}

